---
title: "Assignment07"
output: html_document
date: "2023-06-08"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(rtracklayer)
library(dplyr)
library(ggplot2)
library(purrr)
library(tidyr)
library(readr)
library(tibble)
gtf <- rtracklayer::import('Saccharomyces_cerevisiae.R64-1-1.75.gtf')
gtf_df=as.data.frame(gtf)
```

## Task 1: Get familiar with tidyverse

## Excersise collection: Life expectancy 
Vorwarnung: der code von der Website ging nicht, deswegen wird hoffentlich dieselbe Data benutzt (danke an Sofia )

```{r echo=TRUE}
gapminder_raw <- read.csv("https://raw.githubusercontent.com/resbaz/r-novice-gapminder-files/master/data/gapminder-FiveYearData.csv")
#einlesen und ersten Einträge anschauen
head(gapminder_raw)

```

## Research questions
How did life expectancy change in the course of the last decades ?
Did it change differently between the continents ?
How does life expectancy differs today between the continents ?
--> Generally, life expectancy has been increasing globally in the past decades 
--> In more "developed" countries with better hygiene and medicine life expectancy is higher 
--> in those countries, life expectancy has increased more rapidly 
Source: https://www.statista.com/statistics/270861/life-expectancy-by-continent/

Is life expectancy related to GDP (Gross Domestic Product)? If so, to what degree (and form)? Is this assocication moderated by continent?
--> in general: positive correlation, indicating that higher GDP levels tend to be associated with longer life expectancy
--> but this can vary across continents (regional factors and disparities)

## First steps- loading needed library
```{r echo=TRUE}
#install.packages("tidyverse")
library(tidyverse)
```
## 6 Exercises
## 6.1 Data Wrangling
1. Filter the data for the Americas in 2007, deselect all other variables
```{r echo=TRUE}
# Piping: "and then" -> more functions at one
#first filter the needed rows 
gapminder_raw %>% filter(continent == "Americas", year=="2007") %>% 
#select and show only countries in Americas continent from 2007
select(country, lifeExp)

```
2. Create the variable gdp, defined as the product of population size and gdp per person
```{r echo=TRUE}
# mit allen 
gdp <- gapminder_raw$pop * gapminder_raw$gdpPercap
head(gdp)
```
3. Identify the observation with lowest gdp per year
```{r echo=TRUE}
lowest_gdp<- gapminder_raw %>% filter(gdpPercap==min(gdpPercap))
lowest_gdp
```
4. Identify all observations with above average life expectancy, stratified for each continent
--> also für jeden Kontinent einzeln?
```{r echo=TRUE}
avg_life_exp<- gapminder_raw %>% group_by(continent) %>% summarize(avg_life_exp = mean(lifeExp))
data_con_average <- gapminder_raw %>% left_join(avg_life_exp, by = "continent")
above_average<- data_con_average%>% filter(lifeExp> avg_life_exp)
head(above_average)

```
--> nicht ganz sicher bei der Aufgabe
-> fast die Hälfte an ürsprünglichen Reihen

5. Count the observations identified in the last step
-> count rows
```{r echo=TRUE}
observations <- nrow(above_average)
head(observations)
```

6. Compute the mean life expectancy (grand mean)
```{r echo=TRUE}
grand_mean <- mean(gapminder_raw$lifeExp)
head(grand_mean)
```

7. Compute the mean life expectancy for each year
```{r echo=TRUE}
mean_years <- gapminder_raw %>% group_by(year) %>% summarise(mean_years=mean(lifeExp))
head(mean_years)
```


## 6.2 Data Visualization
1. Create a scatter plot showing the association of gdp per person and life expectancy
```{r echo=TRUE}
library(ggplot2)
ggplot(gapminder_raw, aes(x= gdpPercap, y=lifeExp)) + geom_point() + labs(x= "GDP per person", y="life expectancy") + ggtitle("association between gdp per person and life expectancy")
```

2. Add a rolling average line (LOESS smoother)
```{r echo=TRUE}
ggplot(gapminder_raw, aes(x= gdpPercap, y=lifeExp)) + geom_point() + geom_smooth(method= "loess", se=FALSE, color="blue") + labs(x= "GDP per person", y="life expectancy") + ggtitle("updated association between gdp per person and life expectancy")
```

3. Add a linear model line
```{r echo=TRUE}
ggplot(gapminder_raw, aes(x= gdpPercap, y=lifeExp)) + geom_point() + geom_smooth(method="loess", se=FALSE, color="blue") + geom_smooth(method= "lm", se=FALSE, color="red") + labs(x= "GDP per person", y="life expectancy") + ggtitle("new updated association between gdp per person and life expectancy")
```

4. Create scatter plot
```{r echo=TRUE}

ggplot(mean_years, aes(x=year, y=mean_years)) + geom_point()+geom_line()+labs(x="Year", y="Average life expectancy")+ ggtitle("Average life expectancy by year")
```

5. Modify the last plot so that there is a line for each continent
```{r echo=TRUE}
mean_years_continent <- gapminder_raw %>% group_by(year, continent) %>% summarize(mean_years =mean(lifeExp))

ggplot(mean_years_continent, aes(x=year, y=mean_years, color=continent, group=continent))+ geom_point() + geom_line() +labs(x="year", y="Average life expectancy", color="Continent")+ ggtitle("average life expectancy by year, plus continent ")
```

6. Create a scatter plot showing the association of gdp per person and life expectancy
```{r echo=TRUE}
ggplot(gapminder_raw, aes(x=gdpPercap, y=lifeExp, color=continent)) + geom_point() + labs(x= "GDP", y="Life expectancy", color="Continent") + ggtitle("association of gdp per person and life expectancy")
```

7. Modify last plot so that the size of the dots represents the population size. in addition, increase the transparency of the dots in order to mitigate overplotting
```{r echo=TRUE}
ggplot(gapminder_raw, aes(x = gdpPercap, y = lifeExp, color = continent, size = pop, alpha = 0.5)) +
  geom_point() +
  labs(x = "GDP per person", y = "Life Expectancy", color = "Continent", size = "Population") +
  ggtitle("Association of GDP per person and Life Expectancy") +
  guides(size = "none")
```

8. Modify last plot so that there is a facet for each continent
-> using facet wrap
```{r echo=TRUE}
ggplot(gapminder_raw, aes(x = gdpPercap, y = lifeExp, color = continent, size = pop, alpha = 0.5)) +
  geom_point() +
  labs(x = "GDP per person", y = "Life Expectancy", color = "Continent", size = "Population") +
  ggtitle("Association of GDP per person and Life Expectancy") +
  facet_wrap(~ continent, nrow = 2) +
  guides(size = "none")
```

9. Modify the last plot so that GDP is log transformed
```{r echo=TRUE}
ggplot(gapminder_raw, aes(x = log10(gdpPercap), y = lifeExp, color = continent, size = pop, alpha = 0.5)) +
  geom_point() +
  labs(x = "Log GDP per person", y = "Life Expectancy", color = "Continent", size = "Population") +
  ggtitle("Association of Log GDP per person and Life Expectancy") +
  facet_wrap(~ continent, nrow = 2) +
  guides(size = "none")
```



## Task 2: Recap 1: Yeast GTF

## Give information about how many chromosome, features and subfeature entries are given in the gtf-file. How many entries of each type are in the gtf?
counts of chromosome, feature, subfeature and type entries:
```{r echo=TRUE}
gtf_df %>%
  count(seqnames)

gtf_df %>%
  count(gene_biotype)

gtf_df %>%
  count(gene_id) %>% #not sure what subfeature means here.
  head

gtf_df %>%
  count(type)
```
Is the amount of features on the ’+’-strand the same as on the ’-’-strand? Is the
amount the same on each chromosome?
The amount of features is not the same. And the amount on each chromosome is also different.
```{r echo=TRUE}
gtf_df %>%
  group_by(seqnames) %>%
  count(strand) %>%
  ungroup() 
```
Plot the length distribution of each feature (gene, cds, start-codon,...)

```{r echo=TRUE}
gtf_df %>%
  ggplot() +
  geom_histogram(aes(x = width), binwidth = 1000) +
  facet_wrap(~type, ncol = 2)
```

Plot the relationship of each subfeature to the feature. E.g. how many transcripts
are there per gene, are there transcripts with more than one exon?

Answer: There is only one transcript per gene and there are transcripts with more than one exon.
```{r echo=TRUE}
gtf_df %>%
  filter(type == type) %>%
  group_by(gene_id) %>%
  count(type) %>%
  ungroup() %>%
  as.data.frame() %>%
  ggplot() +
  geom_histogram(aes(x = n), binwidth = 1) +
  facet_wrap(~type, ncol = 2)
```