---
title: "TidyverseRecap"
author: "Anna Lisa und Wenjia"
date: "2023-11-13"
output:
pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggplot2)
gapminder_raw <- read.csv("https://raw.githubusercontent.com/resbaz/r-novice-gapminder-files/master/data/gapminder-FiveYearData.csv")
YEASTRACT <- read.delim("yeastract2019-flat-file.tsv")
```

```{r echo=TRUE}
#einlesen und ersten Einträge anschauen
head(gapminder_raw)
```

Vorweg: gute Wiederholung für tidyverse, unten noch ein paar Beispiele an unseren Data-Sets
## Research questions
How did life expectancy change in the course of the last decades ?
Did it change differently between the continents ?
How does life expectancy differs today between the continents ?
--> Generally, life expectancy has been increasing globally in the past decades 
--> In more "developed" countries with better hygiene and medicine life expectancy is higher 
--> in those countries, life expectancy has increased more rapidly 

## 6.1 Data Wrangling
## Filter the data for the Americas in 2007, deselect all other variables.
```{r echo=TRUE}
library(tidyverse)
#dim(gapminder_raw)
gapminder1 <- gapminder_raw %>% 
  filter(continent == "Americas", year == "2007")%>% 
  select(country, lifeExp)
head(gapminder1)
```


## Create the variable gdp, defined as the product of population size and gdp per person
```{r echo=TRUE}
gapminder_raw <- gapminder_raw %>% 
  mutate(gdp = gdpPercap * pop)
head(gapminder_raw)
```

## Identify the observation with lowest gdp per person.
```{r echo=TRUE}
lowest_gdp<- gapminder_raw %>% filter(gdpPercap==min(gdpPercap))
lowest_gdp
```

## Identify all observations with above average life expectancy, stratified for each continent.
```{r echo=TRUE}
avg_life_exp<- gapminder_raw %>% group_by(continent) %>% summarize(avg_life_exp = mean(lifeExp))
data_con_average <- gapminder_raw %>% left_join(avg_life_exp, by = "continent")
above_average<- data_con_average%>% filter(lifeExp> avg_life_exp)
head(above_average)
```

## Count the observations identified in the last step.
```{r echo=TRUE}
observations <- nrow(above_average) # nrow generell ganz hilfreich
observations
```

## Compute the mean life expectancy (the grand mean; ie., across all observations)
```{r echo=TRUE}
across_mean <- mean(gapminder_raw$lifeExp)
across_mean
```

## Compute the mean life expectancy for each year
```{r echo=TRUE}
mean_years <- gapminder_raw %>% group_by(year) %>% summarise(mean_years=mean(lifeExp))
mean_years
```

## 6.2 Data Visualization
## Create a scatter plot showing the association of gdp per person and life expectancy. Put the putative cause on the X axis and the putative effect on the y axis.
```{r echo=TRUE}
library(ggplot2)
ggplot(gapminder_raw, aes(x= gdpPercap, y=lifeExp)) + geom_point(color="pink") + labs(x= "GDP per person", y="life expectancy") + ggtitle("association between gdp per person and life expectancy")

#in this case we could use log(GDP per person) instead of GDP per person
ggplot(gapminder_raw, aes(x= log(gdpPercap+1), y=lifeExp)) + geom_point(color="blue") + labs(x= "GDP per person", y="life expectancy") + ggtitle("association between gdp per person with log and life expectancy")
```

## Add a rolling average line (also known as LOESS smoother).
```{r echo=TRUE}
ggplot(gapminder_raw, aes(x= gdpPercap, y=lifeExp)) + geom_point() + geom_smooth(method= "loess", se=FALSE, color="red") + labs(x= "GDP per person", y="life expectancy") + ggtitle("updated association between gdp per person and life expectancy")
```

3. Add a linear model line
```{r echo=TRUE}
ggplot(gapminder_raw, aes(x= gdpPercap, y=lifeExp)) + geom_point() + geom_smooth(method="loess", se=FALSE, color="red") + geom_smooth(method= "lm", se=FALSE, color="green") + labs(x= "GDP per person", y="life expectancy") + ggtitle("new updated association between gdp per person and life expectancy")
```

4. Create scatter plot
```{r echo=TRUE}
ggplot(mean_years, aes(x=year, y=mean_years)) + geom_point()+geom_line()+labs(x="Year", y="Average life expectancy")+ ggtitle("Average life expectancy by year")
```

5. Modify the last plot so that there is a line for each continent
```{r echo=TRUE}
mean_years_continent <- gapminder_raw %>% group_by(year, continent) %>% summarize(mean_years =mean(lifeExp))

ggplot(mean_years_continent, aes(x=year, y=mean_years, color=continent, group=continent))+ geom_point() + geom_line() +labs(x="year", y="Average life expectancy", color="Continent")+ ggtitle("average life expectancy by year, plus continent ")
## group und continent für Projekt merken
```

6. Create a scatter plot showing the association of gdp per person and life expectancy
```{r echo=TRUE}
ggplot(gapminder_raw, aes(x=gdpPercap, y=lifeExp, color=continent)) + geom_point() + labs(x= "GDP", y="Life expectancy", color="Continent") + ggtitle("association of gdp per person and life expectancy")

ggplot(gapminder_raw, aes(x=log(gdpPercap), y=lifeExp, color=continent)) + geom_point() + labs(x= "log(GDP)", y="Life expectancy", color="Continent") + ggtitle("association of gdp per person and life expectancy with log")
```

7. Modify last plot so that the size of the dots represents the population size. in addition, increase the transparency of the dots in order to mitigate overplotting
```{r echo=TRUE}
ggplot(gapminder_raw, aes(x = gdpPercap, y = lifeExp, color = continent, size = pop, alpha = 0.5)) +
  geom_point() +
  labs(x = "GDP per person", y = "Life Expectancy", color = "Continent", size = "Population") +
  ggtitle("Association of GDP per person and Life Expectancy") +
  guides(size = "none")
```

8. Modify last plot so that there is a facet for each continent
-> using facet wrap
```{r echo=TRUE}
ggplot(gapminder_raw, aes(x = gdpPercap, y = lifeExp, color = continent, size = pop, alpha = 0.5)) +
  geom_point() +
  labs(x = "GDP per person", y = "Life Expectancy", color = "Continent", size = "Population") +
  ggtitle("Association of GDP per person and Life Expectancy") +
  facet_wrap(~ continent, nrow = 2) +
  guides(size = "none")
```

9. Modify the last plot so that GDP is log transformed
```{r echo=TRUE}
ggplot(gapminder_raw, aes(x = log10(gdpPercap), y = lifeExp, color = continent, size = pop, alpha = 0.5)) +
  geom_point() +
  labs(x = "Log GDP per person", y = "Life Expectancy", color = "Continent", size = "Population") +
  ggtitle("Association of Log GDP per person and Life Expectancy") +
  facet_wrap(~ continent, nrow = 2) +
  guides(size = "none")
```

## Weitere Beispiele using Yeastract dataset
1. violin plot
```{r echo=TRUE}
new_names <- c( "Regulator systematic", "Regulator standard", "Target systematic", "Target standard", "Id", "Date", "Environment", "Regulation", "Type", "Method")
colnames(YEASTRACT) <- new_names

plot_dat <- YEASTRACT %>%
  group_by(`Target systematic`) %>%
  summarise(num_tf = n_distinct(`Regulator systematic`)) %>%
  ggplot(aes(x = "", y = num_tf)) +  # Leere x-Achse
  geom_violin(fill = "violet", color = "white") +
  ggtitle("Number of Transcription Factors associated with a target gene") +
  xlab("under all conditions") + 
  ylab("numbers of TFs") +
  theme_bw()

plot_dat
```

2. combinition of violin plots, with better view of comparision
```{r echo=TRUE}
counts <- data.frame()
for(item in unique(YEASTRACT$Environment)){
  control <- YEASTRACT %>%
    filter(Environment == item) %>%
    group_by(`Target systematic`) %>%
    summarise(num_tf = n_distinct(`Regulator systematic`)) %>%
    mutate(Environment = item)
  counts <- rbind(counts,control)
}
ggplot(counts, aes(x = Environment, y = log(num_tf), fill = Environment)) +
  geom_violin() +
  ggtitle("Number of TFs associated with a target gene ") +
  xlab("Environments") +  
  ylab("numbers of TFs with log") +
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())

```

3. numbers of TFs in different conditions
```{r echo=TRUE}
YEASTRACT %>%
  group_by(`Environment`) %>%
  ggplot(aes(x = Environment, fill = Type)) +  
  geom_bar() +
  ggtitle("Number of TFs in different conditions") +
  xlab("under all conditions") + 
  ylab("numbers of TFs") +
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust = 0.5))
```
