---
title: "Assignment06"
output: html_document
date: "2023-06-03"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Task 1: Tidyverse

## Part 1: Getting started w/ readr

## Installation and setup - Cheatsheet

-\> already done in Assignment 01 -\> We all use RStudio ;)

## Loading packages

Install tidyverse and loading packages -\> designed for data science
gapminder -\> dataset used as an example in data analysis

```{r echo=TRUE}
install.packages(c("tidyverse", "here", "gapminder"))
install.packages("readr")
install.packages("tidyr")
install.packages("dplyr")
```

```{r echo=TRUE}
library(tidyverse)
```

## Data import

useful tools: -`read_csv`, `write_csv` - `read_tsv`, `write_tsv` -
`read_delim`, `write_delim` - `here::here`

-\> we already used many of those, so no useless examples :)

## Knowing your data: basic data exploration

Dataset details:

AA: Agilent Adult AF: Agilent Fetus BA: BioChain Adult BF: BioChain
Fetus OA: Origene Adult

"str()" - shows the structure of the dataframe

"head()" - shows top

"rows" of dataframe

"glimse()"- info-dense summary of the data (!)

"view()" - view data

colnames() , nrows(), ncol() \<- all already known

data[1:5,7:10]- subsetting dataframe

write_tsv(data[1:100,7:12],"data_txt") \<- saves the data file

## Part 2: Resharping data w/ tidyr

TIDY, when each variable has its own column and each observation has its
own row

## Gather, Spread & Unite, Separate

functions used:

"gather()" -\> gather COLUMNS -\> ROWS

"spread()" -\> spread ROWS-\> COLUMNS

"privot_longer()" -\> gather COLUMNS -\> ROWS

"privot_wider()" -\> spread ROWS -\> COLUMNS

"separate()" -\> separate 1 COLUMN -\> many COLUMNS

"unite()" -\> Unite several COLUMNS -\> 1 COLUMN

gather -\> COLUMNS into ROWS(key-value-pairs): WIDE-\> LONG spread -\>
ROWS into COLUMNS: LONG -\> WIDE

unite -\> unite multiple columns into one separate -\> separate one
column into multiple columns

## Part 3: Data wrangling w/dplyr

-\> helpful for basic data wrangling

useful functions:

"filter()" -\> PICK observations by their values \| ROWS

"select()" -\> PICK variables by their names \| COLUMNS

"mutate()" -\> CREATE new variables w/ functions of existing variables
\| COLUMNS

"transmute()"-\> COMPUTE 1 or more COLUMNS but drop original columns

"arrange()" -\> REORDER the ROWS

"summarize()" -\>COLLAPSE many values to a single SUMMARY
"group_by()"-\> GROUP data into rows with the same value of
variable(COLUMN)

"left_join()", "right_join()", "inner_join()", "outer_join()",
"bind_rows()", "bind_cols()" -\> combining possibilities

## Recap of the functions

Filter -\> return rows with matching conditions

Select -\> select/rename variables/columns by name (e.g. select columns
that math a specific pattern)

Mutate -\> adds new variables & keeps existing ones

Transmute -\> adds new var, drops existing var

Distinct -\> pick unique entries

Arrange -\> arrange rows by variables (e.g. sort by length)

Summarize -\> reduces multiple values down to a single value

group_b -\> combine entries by one or more variables

## advantages of wide and long dataframe format

WIDE: 
-easy to understand 
-compact 
-quick data exploration

LONG: 
-\>when one has a sample name and a sample value 
-flexible for analysis 
-easier data manipulation 
-facilitates data visualization

## Task 2: Gather databases

