---
title: "Assignment11"
author: "Anna Lisa and Wenjia"
date: "2023-07-05"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
survey<- read.csv("basic_yeast_heatshock.csv", header=TRUE)
library(dplyr)
library(tidyverse)
```

##Task 1: MSMB 05

Clustering is a useful technique for understanding complex multivariate
data; it is unsupervised machine learning --\> Clustering is more
general than EM-algorithm, and can be applied to more complex data

What is similar ? Different approaches: Euclidean distance
(d(A,B)=Wurzel((a1-b1)\^2 + ....)), Manhattan Block ( distance takes the
sum of the absolute differences in all coordinates), Maximum, weighted
Euklidean distance, Binary und viele viele mehr

## k-means method:

1.  Starts from a matrix of features measured on a set of observations.

2.  Randomly pick distinct cluster centers out of the observations
    ("seeds").

3.  Assign each of the remaining observation to the group to whose
    center it is the closest.

4.  For each group, choose a new center from the observations in the
    group, such that the sum of the distances of group members to the
    center is minimal; this is called the medoid.

5.  Repeat Steps 3 and 4 until the groups stabilize.

## Density-based clustering

This method looks for regions of high density separated by sparser
regions. It has the advantage of being able to cope with clusters that
are not necessarily convex --\> we know DBSCAN

How does it work? concept of density-reachability: a point q is directly
density-reachable from a point p if it is not further away than a given
threshold ,and if p is surrounded by sufficiently many points such that
one may consider p (and q) be part of a dense region

A cluster is then a subset of points that satisfy the following
properties:

1.  All points within the cluster are mutually density-connected.

2.  If a point is density-connected to any point of the cluster, it is
    part of the cluster as well.

3.  Groups of points must have at least MinPts points to count as a
    cluster.

--\> hierarchal clustering is also possible, self explaining (talked
about in Bioinformatik 2)

## How could clustering methods be applied in the analysis we conducted in the previous assignments?

\--\> in general: find structures and relationships and visualize them!

-   Pattern discovery e.g. in the heatshock

-   Classification and Visualization e.g. in yeastract

-   Data Reduction (reduce complexity)


## Task 2: Clustering
```{r echo=TRUE}
head(survey)
```
```{r echo=TRUE}
data <- survey
data$gene_id <- NULL
rownames(data) = survey$gene_id
data_matrix <- as.matrix(log(data+1))
heatmap(data_matrix)
```

## Top 50 highest expressed genes based on row sums

```{r echo=TRUE}
data$sums <- rowSums(data)
top_fifty <- data %>%
         filter(rank(desc(sums))<=50)
top_fifty$sums = NULL
top_fifty = as.matrix(log(top_fifty+1))
heatmap(top_fifty)
data$sums = NULL
```

## Top 50 changing genes determined by the maximum absolute difference between the means of 10min and 0min and the means of 30min and 10min.
```{r echo=TRUE}
data$dif <- abs(rowMeans(data[1:3]) - rowMeans(data[4:6])) + abs(rowMeans(data[4:6]) - rowMeans(data[7:9]))
top_fifty2 <- data %>%
         filter(rank(desc(dif))<=50)
top_fifty2$dif = NULL
top_fifty2 = as.matrix(log(top_fifty2+1))
heatmap(top_fifty2)
data$dif = NULL

```

## Top 50 non-changing genes identified by the minimum absolute difference between the means of 10min and 0min and the means of 30min and 10min. Exclude genes that have not been measured at all.
```{r echo=TRUE}
measured <- data[rowSums(data[])>0,]
measured$dif <- abs(rowMeans(measured[1:3]) - rowMeans(measured[4:6])) + abs(rowMeans(measured[4:6]) - rowMeans(measured[7:9]))
top_fifty3 <- measured %>%
         filter(rank(desc(dif))>=6404)
top_fifty3$dif = NULL
top_fifty3 = as.matrix(log(top_fifty3+1))
heatmap(top_fifty3)

```

## Top 50 genes that only show changes at the beginning, determined by the maximum absolute difference between the means of 10min and 0min and the minimum absolute difference between the means of 30min and 10min.

```{r echo=TRUE}
data$dif <- abs(rowMeans(data[1:3]) - rowMeans(data[4:6])) - abs(rowMeans(data[4:6]) - rowMeans(data[7:9]))
top_fifty2 <- data %>%
         filter(rank(desc(dif))<=50)
top_fifty2$dif = NULL
top_fifty2 = as.matrix(log(top_fifty2+1))
heatmap(top_fifty2)
data$dif = NULL

```

## Clustering and heatmap using pheatmap

```{r echo=TRUE}
library(pheatmap)
row_clusters <- hclust(dist(data_matrix))
column_clusters <- hclust(dist(t(data_matrix))) 
pheatmap(data_matrix,
         clustering_method = "complete",  # Method used for clustering
         cluster_rows = TRUE,  # Perform row clustering
         cluster_cols = TRUE,  # Perform column clustering
         row_cluster_dist = "euclidean",  # Distance metric for row clustering
         col_cluster_dist = "euclidean",  # Distance metric for column clustering
         row_dendrogram = TRUE,  # Display row dendrogram
         col_dendrogram = TRUE,  # Display column dendrogram
         color = colorRampPalette(c("blue", "white", "red"))(100))  # Color palette for heatmap


```
