---
title: "Assignment10"
author: "Anna Lisa and Wenjia"
date: "2023-06-29"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
YEASTRACT <- read.delim("yeastract2019-flat-file.tsv")
```


## Task 1: Biological networks

```{r echo=TRUE}

head(YEASTRACT)
new_names <- c( "Regulator systematic", "Regulator standard", "Target systematic", "Target standard", "Id", "Date", "Environment", "Regulation", "Type", "Method")
colnames(YEASTRACT) <- new_names
head(YEASTRACT)
```

## 1. How many transcription factors/target genes are included in the file?
```{r echo=TRUE}
regulators <- YEASTRACT$`Regulator systematic`
targets <- YEASTRACT$`Target systematic`

num_transcription_factors <- length(unique(regulators))
num_target_genes <- length(unique(targets))

barplot(c(num_transcription_factors, num_target_genes), 
        names.arg = c("Transcription Factors", "Target Genes"),
        ylab = "Count",
        main = "Number of Transcription Factors and Target Genes")
```


## 2. How are the edges described, and how many of them are redundant?
```{r echo=TRUE}
# Combine regulator and target to create unique edge identifiers
edges <- paste(regulators, targets, sep = "-")

# Count the total number of edges
total_edges <- length(edges)

# Identify duplicate edges
redundant_edges <- sum(duplicated(edges))

# Calculate the number of non-redundant edges
non_redundant_edges <- total_edges - redundant_edges

# Print the results
print(paste("Total Edges:", total_edges))
print(paste("Redundant Edges:", redundant_edges))
print(paste("Non-Redundant Edges:", non_redundant_edges))

edge_data <- data.frame(Type = c("Non-Redundant Edges", "Redundant Edges"),
                        Count = c(non_redundant_edges, redundant_edges))

# Create the ggplot
ggplot(edge_data, aes(x = Type, y = Count, fill = Type)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Number of Non-Redundant and Redundant Edges",
       x = "Edge Type",
       y = "Count") +
  theme_minimal()
```


## 3. How many targets are associated with each transcription factor? How many transcription factors are associated with each target?
```{r echo=TRUE}

YEASTRACT %>%
  group_by(`Regulator systematic`) %>%
  summarize(targets = n_distinct(`Target systematic`)) %>%
  head()

YEASTRACT %>%
  group_by(`Regulator systematic`) %>%
  summarize(targets = n_distinct(`Target systematic`)) %>%
  ungroup() %>%
  as.data.frame() %>%
  ggplot() +
  geom_histogram(aes(x = targets), fill = "pink", colour = "black") +
  ggtitle("counts of targets associated with each transcription factor") +
  ylab("frequency")

YEASTRACT %>%
  group_by(`Target systematic`) %>%
  summarize(transcription_factors = n_distinct(`Regulator systematic`)) %>%
  head()

YEASTRACT %>%
  group_by(`Target systematic`) %>%
  summarize(transcription_factors = n_distinct(`Regulator systematic`)) %>%
  ungroup() %>%
  as.data.frame() %>%
  ggplot() +
  geom_histogram(aes(x = transcription_factors), fill = "pink", colour = "black") +
  ggtitle("counts of transcription factors associated with each target") + 
  ylab("frequency")

```
## 4.  What is the timeframe of the data? Show a histogram.

```{r echo=TRUE}
dates = YEASTRACT$Date
dates <- as.Date(dates, format = c("%Y"))
dates <- as.data.frame(dates)
ggplot(dates, aes(x=dates)) + geom_histogram(fill = "pink", colour = "black")
```

## 5. Display the distribution of the columns ’Environment’, ’Regulation’, ’Type’, and ’Method’. Research these attributes on the internet to find more information about their values.

```{r echo=TRUE}

environment <- 
  YEASTRACT %>%
  group_by(Environment) %>%
  summarise(count = n())

ggplot(environment, aes(Environment, count, fill = Environment)) + 
  geom_bar(stat = 'identity')

regulation <- 
  YEASTRACT %>%
  group_by(Regulation) %>%
  summarise(count = n())

ggplot(regulation, aes(Regulation, count)) +   
    geom_bar(stat = 'identity')
  
type <- 
  YEASTRACT %>%
  group_by(Type) %>%
  summarise(count = n())

ggplot(type, aes(Type, count)) +   
    geom_bar(stat = 'identity')

method <- 
  YEASTRACT %>%
  group_by(Method) %>%
  summarise(count = n())

ggplot(method, aes(Method, count)) +   
    geom_bar(stat = 'identity')
```
## 6. Utilize facets on the above columns for the plots. Are there any imbalances?
```{r echo=TRUE}

```

## MSMB 04

EM often used in motif discovery, multiple sequence alignment etc. \##
Biased coinflip Example as known from EiB 2:

```{r echo=TRUE}
coinflips = (runif(10000) > 0.5)
table(coinflips)

oneFlip = function(fl, mean1 = 1, mean2 = 3, sd1 = 0.5, sd2 = 0.5) {
  if (fl) {
   rnorm(1, mean1, sd1)
  } else {
   rnorm(1, mean2, sd2)
  }
}
fairmix = vapply(coinflips, oneFlip, numeric(1))
library("ggplot2")
library("dplyr")
ggplot(tibble(value = fairmix), aes(x = value)) +
     geom_histogram(fill = "purple", binwidth = 0.1)

```

## Discovering the hidden class labels

E step: Assuming that 0 (i.e., the means, standard deviations and) is
known, evaluate the membership weights

M step: Given the membership weights of each observation xi, determine a
new, improved maximum likelihood estimate of 0

## What does the EM-Algorithm do ?

The Expectation-Maximization (EM) algorithm is a statistical method used
to estimate parameters in models with unobserved or missing data. It is
particularly useful when dealing with problems involving incomplete or
latent variables. The algorithm iteratively maximizes a likelihood
function to estimate the parameters of interest

## What mathematical background is used

1.  Maximum Likelihood Estimation (MLE) --\> explained in our
    presentation a few weeks ago

    -MLE is a method for estimating the parameters of a statistical
    model by maximizing the likelihood function. The likelihood function
    measures the probability of observing the given data as a function
    of the model parameters.

    -The goal of MLE is to find the parameter values that maximize the
    likelihood function, making the observed data most likely under the
    assumed model

2.  Latent Variables

    -Latent variables are unobserved or hidden variables in a
    statistical model. They are not directly measured or observed but
    are inferred from other variables.

    -Latent variables are often used to represent underlying concepts or
    processes that are not directly measurable. For example, in
    clustering problems, the latent variables represent the unknown
    cluster assignments of data points.

3.  It is used with Hidden Markov models, where the unobserved data
    would be the states within the underlying Markov chain

## EM algorithm step by step: 

## Step 1: Initialization

```         
-   Start with initial guesses for the model parameters.
-   These initial parameter values can be randomly chosen or based
    on some prior knowledge.
```

Step 2: Expectation Step (E):

\- algorithm computes the expected values of the latent variables given
the observed data and the current parameter estimates

\- calculates the probabilities or posterior distributions of the latent
variables using the current parameter values and the observed data

Step 3: Maximization Step (M):

\- algorithm updates the parameter estimates by maximizing the expected
log-likelihood (ELL) computed in the E-step

\- finds the parameter values that maximize the ELL, treating the
expected values of the latent variables as complete data

Step 4: Iteration:

\- Repeat the E-step and M-step until convergence is reached

## Discuss one example of EM-algorithm:

From EiB 2 --\> On closer inspection, you notice that the GC content of
the sequences varies greatly. In reviewing their experiment, you find
that the Petri dish was contaminated by another bacterium. Now you do
not know from which bacterium the measured sequences originate. Use the
EM algorithm to estimate the maximum likelihoods A and B of both
genomes. Choose starting value 𝜃 A = 0.4 and 𝜃B = 0.6. Since the growth
rate of both bacterial strains is approximately identical, you can
assume that cells of both bacterial species occur equally frequently in
the Petri dish.

Note: In the expectation step, the conditional probability P (A\|k) for
a sequence of bacterium A is calculated using k nucleotides from {G, C}.
The probability of P (k\|A) can be calculated here using the binomial
distribution B (𝜃, n, k).

```{r echo=TRUE}
# E-STEP
calculate_conditional_probabilities <- function(sequences, thetaA, thetaB) {
  conditional_probs <- list()

  for (sequence in sequences) {
    n <- nchar(sequence)
    k <- sum(strsplit(sequence, "")[[1]] %in% c('G', 'C'))

    PA_k <- dbinom(k, n, thetaA)
    PB_k <- dbinom(k, n, thetaB)

    conditional_probs[[sequence]] <- list(PA_k = PA_k, PB_k = PB_k)
  }

  return(conditional_probs)
}

# M-STEP
# --> multiplying the k with the conditional probabilities and dividing the sum by(sum total counts multiplied with conditional probabilities)

update_maximum_likelihoods <- function(sequences, conditional_probs) {
  sumPA_k <- 0
  sumPB_k <- 0
  sumPA_total <- 0
  sumPB_total <- 0

  for (sequence in sequences) {
    n <- nchar(sequence)
    k <- sum(strsplit(sequence, "")[[1]] %in% c('G', 'C'))

    sumPA_k <- sumPA_k + k * conditional_probs[[sequence]]$PA_k
    sumPB_k <- sumPB_k + k * conditional_probs[[sequence]]$PB_k
    sumPA_total <- sumPA_total + n * conditional_probs[[sequence]]$PA_k
    sumPB_total <- sumPB_total + n * conditional_probs[[sequence]]$PB_k
  }

  thetaA <- sumPA_k / sumPA_total
  thetaB <- sumPB_k / sumPB_total

  return(list(thetaA = thetaA, thetaB = thetaB))
}

# Example
sequences <- c("CTGAACGAGG", "CATTCCTTGG", "CGGGCCGGAC", "GTGGACCCTC", "AAAATGTACC")
thetaA <- 0.4
thetaB <- 0.6

# E-step
conditional_probs <- calculate_conditional_probabilities(sequences, thetaA, thetaB)

# M-step
updated_maximum_likelihoods <- update_maximum_likelihoods(sequences, conditional_probs)
thetaA <- updated_maximum_likelihoods$thetaA
thetaB <- updated_maximum_likelihoods$thetaB

cat("Updated Maximum Likelihoods:")
cat("\nTheta A:", thetaA)
cat("\nTheta B:", thetaB)

```
