---
title: "Assignment12"
author: "Anna Lisa and Wenjia"
date: "2023-07-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
survey<- read.csv("basic_yeast_heatshock.csv", header=TRUE)
```

## MSMB 06

## False discovery rate

FDR = FP / (FP+TP)

--\>statistical measure that quantifies the proportion of false discoveries among all the significant results when conducting multiple tests or exploring multiple variables.

\--\> It helps control the risk of drawing incorrect conclusions due to chance variability.

\--\> By controlling the FDR, researchers can ensure that the significant findings they identify are more likely to be true and not just random noise, providing more reliable results in situations with multiple tests or variables.

## p-value

\--\> helps determine the strength of evidence against a null hypothesis

\--\> represents the probability of obtaining results as extreme as, or more extreme than, the observed data, assuming that the null hypothesis is true

## multiple testing

\--\> occurs when performing numerous statistical tests or comparisons

Controlling method: **Bonferroni correction**: (controls Type \| errors)

\--\> adjust alpha, alpha should be the max. significance niveau of all the testings

what should be the max. sig. niv. for each ? --\>

alpha(adjusted) = alpha/ j (j = amount of tests)

```{r echo=TRUE}
library(ggplot2)
library(tidyverse)
m = 10000
ggplot(tibble(
  alpha = seq(0, 7e-6, length.out = 100),
  p     = 1 - (1 - alpha)^m),
  aes(x = alpha, y = p)) +  geom_line() +
  xlab(expression(alpha)) +
  ylab("Prob( no false rejection )") +
  geom_hline(yintercept = 0.05, col = "red")

```

\--\> not the best, but definitely very quick and easy to use

## Null hypothesis / alternative hypothesis 

H0: assumes there is no significant relationship or effect between variables or groups in a statistical analysis

\--\> baseline or default position

H1: opposes the null hypothesis and suggests the presence of a significant relationship or effect between variables or groups.

\--\> observed differences are not due to random chance

Why is the H1 true when the H0 fails ?

\--\> because there must be sufficient evidence to reject the null hypothesis and support the presence of a significant relationship or effect

\--\> one is always true, when the other is being rejected

## Types of error

Type \| error (false positive) --\> rejecting the null hypothesis, although it is true

Type \|\| error (false negative) --\> not rejecting the null hypothesis, although it is false

\--\> reducing one of them, automatically increasing the other one

To keep in mind: specificity and sensitivity



## Task 2: DESeq

```{r echo=TRUE}
library(tidyverse)
library(DESeq2)
library(UpSetR)
library(data.table)

vulcano=function(resultsObject,title){
  par(mar=c(5,5,5,5), cex=1.0, cex.main=1.4, cex.axis=1.4, cex.lab=1.4)
  
  with(resultsObject, plot(log2FoldChange, -log10(padj), pch=20, main=paste("Volcano plot\n",title), cex=1.0, xlab=bquote(~Log[2]~fold~change), ylab=bquote(~-log[10]~Q~adj~value)))
  with(subset(resultsObject, padj<0.05 & abs(log2FoldChange)>1), points(log2FoldChange, -log10(padj), pch=20, col="red", cex=0.5))
  

  abline(v=0, col="black", lty=3, lwd=1.0)
  abline(v=-1, col="black", lty=4, lwd=2.0)
  abline(v=1, col="black", lty=4, lwd=2.0)
  abline(h=-log10(0.05), col="black", lty=4, lwd=2.0)
}

#reform the data
df <- data.frame(survey$gene_id, rowMeans(survey[2:4]), rowMeans(survey[5:7]), rowMeans(survey[8:10]))
colnames(df) <- c("gene_id", "X25_0", "X42_10", "X42_30")
X25_0 <- data.frame(gene_id = df$gene_id, expressions = df$X25_0, condition = factor(rep("X25_0")))
X42_10 <- data.frame(gene_id = df$gene_id, expressions = df$X42_10, condition = factor(rep("X42_10")))
X42_30 <- data.frame(gene_id = df$gene_id, expressions = df$X42_30, condition = factor(rep("X42_30")))
compare1 <- rbind(X25_0, X42_10)
compare2 <- rbind(X25_0, X42_30)

# X25_0 VS X42_10
cts <- cts[, rownames(compare1)]

dds <- DESeqDataSetFromMatrix(countData = cts,
                                colData = compare1,
                                design= ~ condition)
dds <- DESeq(dds)
res <- results(dds)

vulcano(res, "X25_0 VS X42_10")

# X25_0 VS X42_30

cts <- cts[, rownames(compare2)]

dds <- DESeqDataSetFromMatrix(countData = cts,
                                colData = compare1,
                                design= ~ condition)
dds <- DESeq(dds)
res <- results(dds)

vulcano(res, "X25_0 VS X42_30")

```
